---
title: "Data Analysis"
output: html_document
date: "2025-04-17"
---
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(dunn.test)
```

T-tests compare the variation of a continuous variable across two levels of a categorical (nominal and ordinal) variable.
```{r}
#3b: Does PFG differ based on Gender? What is the null hypothesis? What is the alternative hypothesis?
#null hypothesis = Gender does not affect PFG
#alternative hypothesis = Gender affects PFG 
#1. Find the descriptive statistics (mean, sd, and counts) of PFG by Gender.
NHANES2017_18%>%
  drop_na(A1CLevel)%>%
  group_by(Gender)%>%
  summarize(mean(A1CLevel), sd(A1CLevel), count=n())
#2. Run a test to see if the variances of the continuous variable are equal across different levels of the categorical variable:
#var.test(Contvar~Catvar, data=, alternative="two.sided")
var.test(A1CLevel~Gender, data=NHANES2017_18, alternative="two.sided")
#3. We will run the T-test with the following code:
#t.test(Contvar~Catvar, var.equal=FALSE, data=)
t.test(A1CLevel~Gender, var.equal=TRUE, data=NHANES2017_18)
ggplot(NHANES2017_18, aes (x = A1CLevel, y = Gender)) + geom_boxplot(fill = "lavender") + labs(x = "A1C Level", y = "Gender", title = "Mean A1C Level by Gender")
```
The mean A1C Level of Males versus Females is not significantly different. Not expected.

Chi-Squared (categorical variables)
proportion of respondents versus population
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
```
3. Let's examine the relationship between PFGcat and Race:
```{r}
#table_name <- xtabs(~ CatVar1 + CatVar2, data=)
#print the table
#show the table in percentages: prop.table(table_name)*100
A1Ccat_Race_table <- xtabs(~ A1Ccat + Race, data=NHANES2017_18)
A1Ccat_Race_table
prop.table(A1Ccat_Race_table)*100
#b. run the chi-square test:
#chisq.test(table_name)
chisq.test(A1Ccat_Race_table)
#barplot 
ggplot(NHANES2017_18, aes(x = A1Ccat, fill = Race)) + geom_bar(position = "dodge") + labs(title = "A1C Categories by Race", x = "A1C Categories", y = "A1C Level")
```
there is not  a significant relationship betwen A1C level and Race. not expected.

Linear regression model -- what is the most important factor influencing diabetes
predict the outcome of a continuous dependent variable from a continuous independent variable.
can we predict BMI from A1C
```{r}
#model_name <- lm(DepVar~IndVar, data=)
A1C_BMI_LRM <- lm (A1CLevel~BMI, data = NHANES2017_18)
#summary(model_name)
summary(A1C_BMI_LRM)
ggplot(NHANES2017_18, aes(x = BMI, y = A1CLevel)) + geom_point(color = "pink") +
  geom_smooth(method = "lm", se=FALSE, color = "skyblue") + geom_smooth(method = "loess", se = FALSE, color = "purple", linetype = "dashed") + labs(title = "BMI by A1C", x = "BMI", y = "A1C")
```
there is not a significant relationship between BMI and A1C level. not expected

Correlation test
A1C & PFG = should have a positive association
```{r}
library(ggpubr)
```

```{r}
#scatter plot code: #ggplot(aes(x=DepVar, y=IndVar), data=) + geom_point(color="") + geom_smooth(method=lm, color="") + labs(x="", y="") + theme_bw()
NHANES2017_18 %>% ggplot(aes(x=A1CLevel, y=PFG)) + geom_point(color="skyblue") + geom_smooth(method=lm, color="pink") + labs(x="A1C Level", y="Plasma Fasting Glucose") + theme_bw()
#correlation test code: cor.test(data$DepVar, data$IndVar, method="pearson")
cor.test(NHANES2017_18$A1CLevel, NHANES2017_18$PFG, method="pearson")
```
there is a significant relationship between PFG and A1C. This was expected.


K-W test: shows if there is a significant difference in the means and variance of a continuous variable across three or more categorical levels when the continuous variable is not normally distributed
Dunn test:identifies the specific bivariate comparisons that significantly differ from each other
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(dunn.test)
```

```{r}
#4a. Does PFG differ by HHIncome? What is the null hypothesis? What is the alternative hypothesis?
#1. Generate summary statistics of PFG by HHIncome?
#NHANESdata%>%
  #drop_na(HHIncome)%>%
  #summarize(mean(HHIncome), sd(HHIncome), count=n())
#2. Generate a boxplot of PFG by HHIncome to visualize the summary statistics
ggplot(NHANES2017_18, aes(x=HHIncome, y=A1CLevel)) +
  geom_boxplot(fill="lavender") +
  labs(x="Household Income",
       y="A1C Level",
       title = "A1C Level by Household Income")
#3. Run the Kruskal-Wallis test and the post-hoc Dunn test to see if there are any significant differences in PFG by HHIncome. The code for the Kruskal-Wallis and Dunn test is: dunn.test(x=data$contvar, g=data$catvar, kw=TRUE)
dunn.test(x=NHANES2017_18$A1CLevel, g=NHANES2017_18$HHIncome, kw=TRUE)
```
There is not a significant relationship between A1C and Household Income



Odds Ratio
Odds of having Diabetes with health insurance versus without
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(epitools)
library(ggmosaic)
```


```{r}
A1Ccat_HealthCover_odds_table <- table(NHANES2017_18$A1Ccat, NHANES2017_18$HealthCovered)
A1Ccat_HealthCover_odds_table
oddsratio(A1Ccat_HealthCover_odds_table)
ggplot(data=NHANES2017_18) + geom_mosaic(aes(x=product(A1Ccat), fill=HealthCovered), na.rm=TRUE, show.legend=FALSE) + labs(title="A1C Cacategories by Health Insurance Coverage", x="A1C Categories", y="Health Insurance Coverage")
```
No statistically significant relationship between A1C level and Health Insurance Coverage. not expacted.

